---
title: "earthquakr: Visualizing NOAA's Significant Earthquakes"
author: "Tyler Byers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{About earthquakr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This is a *draft* vignette for the `earthquakr` package, drafted for the mid-term status check due on April 23, 2017.  Expect a more robust vignette for the final project due by May 7.

## Installing earthquakr

To install the earthquakr, the user will need to use the `devtools::install_github` function. More details will be provided once package development is complete.

## Necessary Packages

```{r echo=FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 5,
                      fig.align = 'center')
```

To work through these examples, the following packages need to be installed and loaded:
```{r message = FALSE}
library(earthquakr) 
library(dplyr)
library(ggplot2)
library(readr)
library(lubridate)
```

## Loading Data

The NOAA Significant Earthquake dataset is provided with this package, using the data set name `quakes`.  The data for this package were pulled on April 1, 2017.  If you wish to use data updated after this date, please see the dataset documentation (`?earthquakr:quakes`) for source.

### Load and Clean Package Data

To load the `quakes` data, simply use the `earthquakr::quakes` command (the `earthquakr::` invocation is necessary because R comes loaded with another data set called `quakes`).

Then call the `eq_clean_data` and `eq_location` clean command to "clean" some of the variables in the data set for use with the visualization tools in this package.  The usage of these commands, plus the "tail" of the data set, are shown below.

```{r}
quakes <- earthquakr::quakes # loads quakes data with data set
quakes <- quakes %>%
  eq_clean_data() %>%
  eq_location_clean()
tail(quakes)
```


### Single Step, Using Package Data

To save a couple steps of command typing, the `eq_load_clean_data` command will load the `quakes` data and do the cleaning steps all with a single command call:

```{r}
quakes <- eq_load_clean_data()
tail(quakes)
```

### Load and Clean Downloaded Data

If you wish to use NOAA Significant Earthquake data updated after April 1, 2017, please visit the NOAA Significant Earthquakes site at  https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1 and download the data to your working directory. Then you may load and clean data for analysis using the following sequence of commands (replace `filename` with the location of your file):

```{r message = FALSE}
filename <- system.file('extdata', 'earthquakes.txt', package = 'earthquakr')
quakes_from_raw <- readr::read_delim(filename, delim = '\t')
quakes_from_raw_clean <- quakes_from_raw %>%
  eq_clean_data() %>%
  eq_location_clean()
tail(quakes_from_raw_clean)
```


## Plot Timeline of Data

This package includes a "timeline" capability to visualize countries' significant earthquakes.

More to be added to this section's documentation...

### Timeline Geoms

Load clean data to be used for all charts below:

```{r}
quakes <- eq_load_clean_data()
```

Basic timeline geom, one country:

```{r}
quakes %>%
  dplyr::filter(COUNTRY == 'USA') %>%
  dplyr::filter(DATE > '2000-01-01') %>%
  ggplot() +
  geom_timeline(aes(x = DATE, y = COUNTRY, color = TOTAL_DEATHS,
                    size = EQ_PRIMARY)) +
  scale_size_continuous(name = 'Richter scale value') +
  scale_color_continuous(name = '# of Deaths')
```

Basic timeline geom, two countries:

```{r}
quakes %>%
  dplyr::filter(COUNTRY %in% c('USA', 'UK')) %>%
  dplyr::filter(DATE > '2000-01-01') %>%
  ggplot() +
  geom_timeline(aes(x = DATE, y = COUNTRY, color = TOTAL_DEATHS,
                    size = EQ_PRIMARY)) +
  scale_size_continuous(name = 'Richter scale value') +
  scale_color_continuous(name = '# of Deaths')
```


### Timeline and Label Geoms

Timeline and label geoms with 2 countries:

```{r}
quakes %>%
  dplyr::filter(COUNTRY %in% c('NEW ZEALAND', 'SOUTH AFRICA')) %>%
  dplyr::filter(DATE > '2000-01-01', DATE < '2015-01-01') %>%
  ggplot() +
  geom_timeline(aes(x = DATE, y = COUNTRY, color = TOTAL_DEATHS,
                    size = EQ_PRIMARY)) +
   geom_timeline_label(aes(x = DATE, y = COUNTRY, magnitude = EQ_PRIMARY,
                         label = LOCATION_NAME, n_max = 5)) +
  scale_size_continuous(name = 'Richter scale value') +
  scale_color_continuous(name = '# of Deaths')
```


### Timeline and Label Geoms with theme_eq

Also added a new `ggplot2` theme: `theme_eq`.  This makes the plots look much nicer and formats them for easier reading:

```{r}
quakes %>%
  dplyr::filter(COUNTRY %in% c('NEW ZEALAND', 'SOUTH AFRICA')) %>%
  dplyr::filter(DATE > '2000-01-01', DATE < '2015-01-01') %>%
  ggplot() +
  geom_timeline(aes(x = DATE, y = COUNTRY, color = TOTAL_DEATHS,
                    size = EQ_PRIMARY)) +
   geom_timeline_label(aes(x = DATE, y = COUNTRY, magnitude = EQ_PRIMARY,
                         label = LOCATION_NAME, n_max = 5)) +
  scale_size_continuous(name = 'Richter scale value') +
  scale_color_continuous(name = '# of Deaths') +
  theme_eq()
```


### Use Wrapper Function for Ease of Use

That is an awful lot of typing to get these charts. So, we created a simple `eq_timeline` wrapper function to make a nice labeled (or not!) chart with the default `aes` values and theme selection.  This is much easier than typing in all of the above.

#### One Country, No Labels

Here is an example with one country and no labels on the earthquakes.

```{r}
quakes %>% eq_timeline(countries = 'NEW ZEALAND', 
                       date_min = as.Date('1995-01-01'),
                       date_max = as.POSIXct('2015-01-01'), 
                       label_n = 0)
```

#### Multiple Countries, Labels

Here is an example of multiple countries and up to 5 labels per country.

```{r}
quakes %>% eq_timeline(countries = c('NEW ZEALAND', 'HAITI'),
                      date_min = '2000-01-01', 
                      date_max = '2015-01-01',
                      label_n = 5)
```


## Plot Map

Finally, this package includes functions for creating an interactive map of the earthquakes using the `leaflet` package.

### Basic Annotations

Call the `eq_map` function with cleaned and filtered data, and specify an annotation column with the argument `annot_col`:

```{r}
quakes %>%
  dplyr::filter(COUNTRY == 'JAPAN') %>%
  dplyr::filter(lubridate::year(DATE) >= 2000) %>%
  eq_map(annot_col = 'DATE')
```

### More Useful Annotations

If you'd like more useful annotations, first call `quakes <- quakes %>% dplyr::mutate(popup_text = eq_create_label(.))` to create a data column with formatted HTML for more useful quakes information.

```{r}
quakes %>%
  dplyr::filter(COUNTRY == 'MEXICO') %>%
  dplyr::filter(lubridate::year(DATE) >= 2000) %>%
  dplyr::mutate(popup_text = eq_create_label(.)) %>%
  eq_map(annot_col = 'popup_text')
```


## Future Developments

For the final assignment due in two weeks, I will clean up this vignette and my code based on any feedback and Travis build requirements.

There will likely be very little additional development on this package.

Future plans for this package include making an R Shiny app with the `flexdashboard` package, and using this package as an underpinning for making interactive plots and timelines, and posting on my blog.
